---
title: "HW 2 exploration"
author: "Hayden Vega"
date: "2/6/25"
---

# HW 2 Exploration

```{r, warning=FALSE, message=FALSE}
# library
library(janitor)
library(tidyverse)
library(ggeffects)
library(sjlabelled)
library(lme4)
library(DHARMa)
library(parameters)
library(effects)
library(MuMIn)
```

#Load in data for each trait with individual ID delimiter

```{r, warning=FALSE, message=FALSE}
height_all <- read.csv("final_height.csv", header = TRUE, sep = ",") %>% separate_wider_delim(specimen_id, delim = "-", names = c("date", "site", "individual", "specimen"), cols_remove=FALSE) %>% 
  unite("individual_id", date:individual, remove = TRUE, sep = "-")

width_all <- read.csv("final_width.csv", header = TRUE, sep = ",") %>% separate_wider_delim(specimen_id, delim = "-", names = c("date", "site", "individual", "specimen"), cols_remove=FALSE) %>% 
  unite("individual_id", date:individual, remove = TRUE, sep = "-")

perimeter_all <- read.csv("final_perimeter.csv", header = TRUE, sep = ",") %>% separate_wider_delim(specimen_id, delim = "-", names = c("date", "site", "individual", "specimen"), cols_remove=FALSE) %>% 
  unite("individual_id", date:individual, remove = TRUE, sep = "-")

surf_all <- read.csv("final_surf.csv", header = TRUE, sep = ",") %>% separate_wider_delim(specimen_id, delim = "-", names = c("date", "site", "individual", "specimen"), cols_remove=FALSE) %>% 
  unite("individual_id", date:individual, remove = TRUE, sep = "-")

tdmc_all <- read.csv("final_tdmc.csv", header = TRUE, sep = ",") %>% separate_wider_delim(specimen_id, delim = "-", names = c("date", "site", "individual", "specimen"), cols_remove=FALSE) %>% 
  unite("individual_id", date:individual, remove = TRUE, sep = "-")

thickness_all <- read.csv("final_thickness.csv", header = TRUE, sep = ",") %>% separate_wider_delim(specimen_id, delim = "-", names = c("date", "site", "individual", "specimen"), cols_remove=FALSE) %>% 
  unite("individual_id", date:individual, remove = TRUE, sep = "-")

volume_all <- read.csv("final_volume.csv", header = TRUE, sep = ",") %>% separate_wider_delim(specimen_id, delim = "-", names = c("date", "site", "individual", "specimen"), cols_remove=FALSE) %>% 
  unite("individual_id", date:individual, remove = TRUE, sep = "-")



```

```{r, warning=FALSE, message=FALSE}
# ggplot theme
theme_set(theme_bw() +
            theme(panel.grid = element_blank()))

# abline color
ref_line_col <- "grey"

# model prediction color
model_col <- "darkblue"
```

#Create multiple regression for each trait - initial & rehydrate WITH & WITHOUT species

```{r, warning=FALSE, message=FALSE}

height_reg_ir <- lmer(height_i ~ height_r * species + (1 | individual_id), data = height_all)
summary(height_reg_ir)

height_reg_ir_nospecies <- lmer(height_i ~ height_r + (1 | individual_id), data = height_all)
summary(height_reg_ir_nospecies)
```

```{r width-models, warning=FALSE, message=FALSE }
width_reg_ir <- lmer(width_i ~ width_r * species + (1 | individual_id), data = width_all)


width_reg_ir_nospecies <- lmer(width_i ~ width_r + (1 | individual_id), data = width_all)
summary(width_reg_ir_nospecies)
```

```{r perimeter-models, warning=FALSE, message=FALSE }
perimeter_reg_ir <- lmer(per_i ~ per_r * species + (1 | individual_id), data = perimeter_all)
summary(perimeter_reg_ir)

perimeter_reg_ir_nospecies <- lmer(per_i ~ per_r + (1 | individual_id), data = perimeter_all)
summary(perimeter_reg_ir_nospecies)
```

```{r surf-models, warning=FALSE, message=FALSE}
surf_reg_ir <- lmer(surf_i ~ surf_r * species + (1 | individual_id), data = surf_all)
summary(surf_reg_ir)

surf_reg_ir_nospecies <- lmer(surf_i ~ surf_r + (1 | individual_id), data = surf_all)
summary(surf_reg_ir_nospecies)
```

```{r tdmc-models, warning=FALSE, message=FALSE}
tdmc_reg_ir <- lmer(weight_i ~ weight_r * species + (1 | individual_id), data = tdmc_all)
summary(tdmc_reg_ir)

tdmc_reg_ir_nospecies <- lmer(weight_i ~ weight_r + (1 | individual_id), data = tdmc_all)
summary(tdmc_reg_ir_nospecies)

```

```{r thick-models, warning=FALSE, message=FALSE}
thick_reg_ir <- lmer(thick_i ~ thick_r * species + (1 | individual_id), data = thickness_all)
summary(thick_reg_ir)

thick_reg_ir_nospecies <- lmer(thick_i ~ thick_r + (1 | individual_id), data = thickness_all)
summary(thick_reg_ir_nospecies)
```

```{r volume-models, warning=FALSE, message=FALSE}
volume_reg_ir <- lmer(vol_i ~ vol_r * species + (1 | individual_id), data = volume_all)
summary(volume_reg_ir)

volume_reg_ir_nospecies <- lmer(vol_i ~ vol_r + (1 | individual_id), data = volume_all)
summary(volume_reg_ir_nospecies)
```

#Compare models using AIC;species & no species for Initial and Rehydrate

```{r height-AIC, warning=FALSE, message=FALSE}
model.sel(height_reg_ir, height_reg_ir_nospecies)
#nospecies is better
```

```{r width-AIC, warning=FALSE, message=FALSE}
model.sel(width_reg_ir, width_reg_ir_nospecies)
#nospecies is better
```

```{r perimeter-AIC, warning=FALSE, message=FALSE}
model.sel(perimeter_reg_ir, perimeter_reg_ir_nospecies)
#with species is better
```

```{r surf-AIC, warning=FALSE, message=FALSE}

model.sel(surf_reg_ir, surf_reg_ir_nospecies)
#species is better
```

```{r tdmc-AIC, warning=FALSE, message=FALSE}
model.sel(tdmc_reg_ir, tdmc_reg_ir_nospecies)
#species is better
```

```{r thick-AIC, warning=FALSE, message=FALSE}
model.sel(thick_reg_ir, thick_reg_ir_nospecies)
#nospecies is better, but doesn't really matter cause it's not a good predictor
```

```{r volume-AIC, warning=FALSE, message=FALSE}
model.sel(volume_reg_ir, volume_reg_ir_nospecies)
#with species is better
```

#Calculating R\^2 Values for the better models

```{r r^2-height, warning=FALSE, message=FALSE}
MuMIn::r.squaredGLMM(height_reg_ir_nospecies)
```

```{r r^2-width, warning=FALSE, message=FALSE}
MuMIn::r.squaredGLMM(width_reg_ir_nospecies)

```

```{r r2-perimeter, warning=FALSE, message=FALSE}
MuMIn::r.squaredGLMM(perimeter_reg_ir)
```

```{r r2-surf, warning=FALSE, message=FALSE}
MuMIn::r.squaredGLMM(surf_reg_ir)

```

```{r r2-tdmc, warning=FALSE, message=FALSE}
MuMIn::r.squaredGLMM(tdmc_reg_ir)
```

```{r r2-thick, warning=FALSE, message=FALSE}
MuMIn::r.squaredGLMM(thick_reg_ir_nospecies)
```

```{r r2-volume, warning=FALSE, message=FALSE}
MuMIn::r.squaredGLMM(volume_reg_ir)
```

#Save rehydration predictions as an object using ggpredict (so we can plot later)

```{r, warning=FALSE, message=FALSE}
height_ir_predictions <- ggpredict(height_reg_ir_nospecies, terms = "height_r[0:45]") %>% 
  mutate(trait = "height")

height_ir_predictions
```

```{r, warning=FALSE, message=FALSE}
width_ir_predictions <- ggpredict(width_reg_ir_nospecies, terms = "width_r[0:30]") %>% #max width = 24
  mutate(trait = "width") 

# look at the data frame
width_ir_predictions

```

```{r, warning=FALSE, message=FALSE}
perimeter_ir_predictions <- ggpredict(perimeter_reg_ir, terms = "per_r[0:375]") %>% 
  mutate(trait = "perimeter")

perimeter_ir_predictions_species <- ggpredict(perimeter_reg_ir, terms = c("per_r", "species")) %>% 
  as_tibble() %>% 
  rename(species = group)
# look at the data frame
perimeter_ir_predictions
```

```{r, warning=FALSE, message=FALSE}
surf_ir_predictions <- ggpredict(surf_reg_ir, terms = "surf_r[0:430]") %>% 
  mutate(trait = "surface area")

surf_ir_predictions_species <- ggpredict(surf_reg_ir, terms = c("surf_r", "species")) %>% 
  as_tibble() %>% 
  rename(species = group)
# look at the data frame
surf_ir_predictions
```

```{r, warning=FALSE, message=FALSE}
tdmc_ir_predictions <- ggpredict(tdmc_reg_ir, terms = "weight_r[0:29050]") %>% 
  as_tibble() %>% 
  mutate(trait = "thallus dry matter content")

tdmc_ir_predictions_species <- ggpredict(tdmc_reg_ir, terms = c("weight_r", "species")) %>% 
  as_tibble() %>% 
  rename(species = group)
# look at the data frame
tdmc_ir_predictions

```

```{r rehydrate_thick - prediction, warning=FALSE, message=FALSE}
thick_ir_predictions <- ggpredict(thick_reg_ir_nospecies, terms = "thick_r[0:2]")  %>% 
  mutate(trait = "thickness")

# look at the data frame
thick_ir_predictions

```

```{r rehydrate_volume - prediction, warning=FALSE, message=FALSE}
volume_ir_predictions <- ggpredict(volume_reg_ir, terms = "vol_r[0:30]") %>% 
  mutate(trait = "volume")

volume_ir_predictions_species <- ggpredict(volume_reg_ir, terms = c("vol_r", "species")) %>% 
  as_tibble() %>% 
  rename(species = group)
# look at the data frame
volume_ir_predictions

```

#Visualize the rehydrate models without species

```{r, warning=FALSE, message=FALSE}
height_ir_plot <- ggplot() +
  
  # plot the raw data as points
  geom_point(data = height_all,
             aes(x = height_r,
                 y = height_i),
             alpha = 0.5,
             color = model_col,
             shape = 21) +
  
  # plot a 1:1 reference line
  geom_abline(slope = 1, 
              intercept = 0,
              linetype = 2,
              linewidth = 1,
              color = ref_line_col) +
  
  # plot the confidence interval 
  geom_ribbon(data = height_ir_predictions,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2,
              fill = model_col) +
  
  # plot the prediction line
  geom_line(data = height_ir_predictions,
            aes(x = x,
                y = predicted),
            color = model_col,
            linewidth = 1) +
  
  # labels
  labs(x = "Rehydrate height (mm)",
       y = "Initial height (mm)") +
  
  # controlling axes to make the plot look square
  # this makes it easier to see the difference between the model and the 1:1 line
  scale_x_continuous(breaks = seq(from = 0, to = 50, by = 10), 
                     limits = c(0, 50)) +
  scale_y_continuous(breaks = seq(from = 0, to = 50, by = 10), 
                     limits = c(0, 50))+
  #add equation + r squared
   annotate("text", x = 10, y = 45, label= "y = 1.02x + 0.0407
R squared = 0.996")
  

height_ir_plot
```

# 2c Questions

1.  I have learned that linear models vary in efficacy based on the incorporation of species.
2.  Yes, but I think I still need to review if my residuals look okay! I might have to restructure data, but I hope not. I'm still trying to think about what I want to say about my data. I want to create a summary figure, but I'm not sure what I should include.
3.  I think I'll have a really hard time with my residual checks and creating a summary figure. I should go to office hours.
